Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2023-05-06 manuscript/manuscript.qmd
Add a bunch of citations

This research note makes two contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding of adherence to derogations. When states comply with derogations, they are actually limiting their obligations to the human rights treaty overall. As such, we should in fact expect a reduction in compliance with the overall human rights treaty during that time. We explore how the potential decline in compliance may actually be a good indication that states are using treaty design as intended.

@@

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021], as part of a broader trend of targeting groups during times of crises [@Grasse2021]. We find that overall, derogating states during the pandemic did not increase abuse and violations. This finding speaks to international legal compliance scholarship and supports the idea that states are being sincere about their intentions and not, generally, using actions like derogation to cover abusive behavior.

@@

We expect that states that derogate during COVID-19 with the intention to impose emergency policies will follow through. Our expectation comes from the specific, time-limited nature of derogations and the limited audience they speak to. Critics of international law argue that states ratify treaties with the expectation of reputational and other benefits. However, derogations have no such public audience and are not signaling an increased commitment to human rights. There is no clear or expected connection with foreign aid, trade, or other theorized 'rewards' for larger legal actions. Instead, derogations are a very specific legal action that states voluntarily enter into after commitment has fully taken place.  Given this, we expect that derogating states took the legal action seriously and generally followed through with the derogations they submitted.

>>>>

This research note makes two contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Cole:2012; @NielsenSimmons:2015; @ZvobgoSandholtzMulesky:2020; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding of adherence to derogations. When states comply with derogations, they are actually limiting their obligations to the human rights treaty overall. As such, we should in fact expect a reduction in compliance with the overall human rights treaty during that time. We explore how the potential decline in compliance may actually be a good indication that states are using treaty design as intended.

@@

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Neumayer:2005; @GoodliffeHawkins:2006; @Simmons2009; @Merry:2009; @Hillebrecht:2014; @ConradRitter:2019]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021], as part of a broader trend of targeting groups during times of crises [@Grasse2021]. We find that overall, derogating states during the pandemic did not increase abuse and violations. This finding speaks to international legal compliance scholarship and supports the idea that states are being sincere about their intentions and not, generally, using actions like derogation to cover abusive behavior.

@@

We expect that states that derogate during COVID-19 with the intention to impose emergency policies will follow through. Our expectation comes from the specific, time-limited nature of derogations and the limited audience they speak to. Critics of international law argue that states ratify treaties with the expectation of reputational and other benefits [@Hathaway:2007; @Hafner-BurtonLeVeckVictor:2017]. However, derogations have no such public audience and are not signaling an increased commitment to human rights. There is no clear or expected connection with foreign aid, trade, or other theorized 'rewards' for larger legal actions. Instead, derogations are a very specific legal action that states voluntarily enter into after commitment has fully taken place. Given this, we expect that derogating states took the legal action seriously and generally followed through with the derogations they submitted.
